
# Residential Mental Health Care (RMHC) Services

```{r, message=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(forcats)
library(scales)
library(gridExtra)
library(cowplot)
library(stringr)
library(kableExtra)

sex_palette <- c("#4CAF50","#F44336","#2196F3")
sex_palette2 <- c("#F44336","#2196F3")
```


## About the Data

**Residential mental health care (RMHC)** services provide specialised mental health care on an overnight basis in a domestic-like environment. RMHC services may include rehabilitation, treatment or extended care.

The data used for this section came from [Australian Institute of Health and Welfare (AIHW)](https://www.aihw.gov.au/mental-health/topic-areas/residential-services).

```{r}
data_dir <- file.path(here::here(), "data", "AIHW")
rmhc3 <- readRDS(file.path(
  data_dir,
  "rmhc3-ppl-accessing-rmhc.rds"
))


# Table name: *`r rmhc3$table_name[1]`*. 
# 
# Table code: *`r rmhc3$table_code[1]`*. 
# 
# Data source: *`r rmhc3$source[1]`*.
```

## Resident Characteristics

*Note: A **resident** is a person who receives residential care intended to be for a minimum of 1 night.*

### By Age Group

```{r}
#| label: fig-rmhc-ra
#| fig-cap: People Accessing RMHC Services by Age Group, 2020-21
rmhc3 %>% 
  filter(demographic == "Age_group") %>% 
  ggplot(aes(x = demographic_variable, y = value, fill = demographic_variable))+
  geom_col()+
  facet_wrap(~measurement,
             scales = "free_y",
             ncol = 1)+
  scale_y_continuous(labels = comma)+
  labs(x = "Age group",
       y = "Value")+
  theme_bw()+
  scale_fill_brewer(palette = "Blues")+
  theme(legend.position = "none")
```

### By Sex

```{r}
#| label: fig-rmhc-rs
#| fig-cap: People Accessing RMHC Services by Sex, 2020-21

rmhc3 %>% 
  filter(demographic == "Sex") %>% 
  ggplot(aes(x = demographic_variable, y = value, fill = demographic_variable))+
  geom_col()+
  facet_wrap(~measurement,
             scales = "free_y",
             ncol = 3)+
  scale_y_continuous(labels = comma)+
  labs(x = "Sex",
       y = "Value")+
  theme_bw()+
  scale_fill_manual(values = sex_palette2)+
  theme(legend.position = "none")
```

### Summary

-   Individuals aged 18-24 have the highest rate of accessing RMHC, at approximately 6 per 10,000 population. For those aged 25-54, the rate stabilizes at around 4 per 10,000.

-   Females have a higher rate, population, and percentage of accessing RMHC.

## RMHC Service Episodes Rate

**Episodes of residential care** are defined as a period of care between the start of residential care (either through the formal start of the residential stay or the start of a new reference period (that is, 1 July)) and the end of residential care (either through the formal end of residential care, commencement of leave intended to be greater than 7 days, or the end of the reference period (that is, 30 June)). An individual can have one or more episodes of care during the reference period.

```{r}
# load the data
data_dir <- file.path(here::here(), "data", "AIHW")
rmhc4 <- readRDS(file.path(
  data_dir,
  "rmhc4-rmhc-by-sex-age-states.rds"
))


# Table name: *`r rmhc4$table_name[1]`*. 
# 
# Table code: *`r rmhc4$table_code[1]`*. 
# 
# Data source: *`r rmhc4$source[1]`*.
```

```{r}
# load the data
data_dir <- file.path(here::here(), "data", "AIHW")
rmhc7 <- readRDS(file.path(
  data_dir,
  "rmhc7-rmhc-ts-by-age-sex.rds"
))

rmhc7_rsa <- rmhc7 %>% 
  filter(count == "Rate_per_10_000_population",
         demographic %in% c("Sex", "Age_group"))
rmhc7_rsa$year <- gsub("–\\d{2}", "", rmhc7_rsa$year)
rmhc7_rsa$year <- as.numeric(rmhc7_rsa$year)

# Table name: *`r rmhc7$table_name[1]`*. 
# 
# Table code: *`r rmhc7$table_code[1]`*. 
# 
# Data source: *`r rmhc7$source[1]`*.

```

### General Sex and Age Characteristics

#### By Sex and Age Group

```{r}
#| label: fig-rmhc-gsa
#| fig-cap: Episodes Rate by Age Group and Sex, 2020-21
sex_palette <- c("#4CAF50","#F44336","#2196F3")

rmhc4 %>% 
  filter(count != "Number",
         states_and_territories == "Total") %>% 
  ggplot(aes(x = age_group, y = value, fill = sex))+
  geom_bar(position="dodge", stat="identity")+
  scale_fill_manual(values = sex_palette)+
  labs(x = "Age group",
       y = "Rate (per 10,000 population)",
       fill = "Sex")+
  theme_bw()
```

#### Summary

-   Females aged 18-24 exhibit an exceptionally high episode rate, exceeding 10 per 10,000 population.

-   For ages 25-54, the episode rates are generally similar across both genders, though slightly higher for females, with an average of about 5 per 10,000 population.

-   Young girls aged 12-17 experience exceptionally higher episode rates than boys in the same age group.

### Sex and Age Characteristics by States or Territories

#### By Sex and States or Territories

```{r}
#| label: fig-rmhc-gss
#| fig-cap: Episodes Rate by Sex and States or Territories, 2020-21
rmhc4 %>% 
  filter(count != "Number",
         #states_and_territories != "Total",
         age_group != "Total") %>% 
  ggplot(aes(x = states_and_territories, y = value, fill = sex))+
  geom_bar(position="dodge", stat="identity")+
  scale_fill_manual(values = sex_palette)+
  labs(x = "States or territories",
       y = "Rate (per 10,000 population)",
       fill = "Sex")+
  theme_bw()
```

*Note: Data for **ACT** is not applicable and should not be interpreted as zero. Some data for **NSW** is rounded to zero.*

#### By Sex, Age Group, States or Territories

```{r}
#| label: fig-rmhc-gsas
#| fig-cap: Episodes Rate by Sex, Age Group, States or Territories, 2020-21
rmhc4 %>% 
  filter(count != "Number",
         age_group != "Total",
         states_and_territories != "Total"
         ) %>% 
  ggplot(aes(x = age_group, y = value, fill = sex))+
  geom_bar(position="dodge", stat="identity")+
  scale_fill_manual(values = sex_palette)+
  facet_wrap(~states_and_territories,
             ncol = 2)+
  labs(x = "Age group",
       y = "Rate (per 10,000 population)",
       fill = "Sex")+
  theme_bw()
```

*Note: Data for **ACT** is not applicable and should not be interpreted as zero. Some data for **NSW** is rounded to zero.*

#### Summary

-   Tasmania has the highest episode rate among all states or territories, while New South Wales registers have the lowest, nearly reaching zero.

-   Female rates are generally higher across most states or territories, notably among young women aged 18-24 in Victoria and South Australia.

-   The Northern Territory is an exception where the male rate is about double the female rate, attributed to a low female rate.

-   Episode rates in Queensland and Western Australia are moderate, with all values less than 10 per 10,000 population.

### Episode Rates Changes Over Time

#### Average Annual Episodes Change

```{r}
## Load data
data_dir <- file.path(here::here(), "data", "AIHW")
rmhc6 <- readRDS(file.path(
  data_dir,
  "rmhc6-rmhc-per-10000-pop-by-sex-n-age.rds"
))

rmhc6$year <- gsub("–\\d{2}", "", rmhc6$year)
rmhc6$year <- as.numeric(rmhc6$year)

# Table name: *`r rmhc6$table_name[1]`*. 
# 
# Table code: *`r rmhc6$table_code[1]`*. 
# 
# Data source: *`r rmhc6$source[1]`*.
```

```{r}
#| label: fig-rmhc-asa
#| fig-cap: Average Annual Episodes Change by Sex and Age group, 2016–17 to 2020–21
sex_palette <- c("#4CAF50","#F44336","#2196F3")

ggplot(rmhc6, 
       aes(x = age_group,
           y = average_annual_change_per_cent_2016_17_to_2020_21,
           fill = sex)) +
  geom_bar(position="dodge", stat="identity")+
  scale_fill_manual(values = sex_palette)+
  labs(x = "Age group",
       y = "Average annual change (%)",
       fill = "Sex")+
  theme_bw()
  
```

#### Episodes Changes Over the Year by Age and Sex

```{r}
#| label: fig-rmhc-cras
#| fig-cap: Change of Episodes Rate by Sex Over Year 2005–06 to 2020–21
rmhc7_rsa %>% 
  filter(demographic == "Sex") %>% 
  ggplot(aes(x = year, y = value, color = demographic_variable)) +
  geom_line()+
  labs(x = "Year",
       y = "Rate (per 10,000 population)",
       color = "Sex")+
  theme_bw()+
  scale_color_manual(values = sex_palette2)

# for analysis
rmhc7_a <- rmhc7_rsa %>% 
  filter(demographic == "Sex",
         year == min(year) | year == max(year)) 
```

```{r}
#| label: fig-rmhc-cnas
#| fig-cap: Change of Episodes Number by Age and Sex Over Year 2005–06 to 2020–21
ggplot(rmhc6, aes(x = year, y = episodes_per_10_000_pop, color = sex)) +
  geom_line()+
  facet_wrap(~age_group)+
  labs(x = "Year",
       y = "Episodes (per 10,000 population)",
       color = "Sex")+
  scale_color_manual(values = sex_palette)+
  theme_bw()
```

### Summary

-   From 2005-06 to 2020-21, the female episode rate quadrupled from 0.9 to 3.9, while the male rate doubled.

-   From 2005 to 2012, the male episode rate was higher than females; between 2012 and 2015, the rates for both genders were the same; from 2015 onwards, the female episode rate surpassed the male rate. This symptom is largely driven by the trend in young women aged 18-24.