
# Specialist Homelessness Services

```{r, message=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(forcats)
library(scales)
library(gridExtra)
library(cowplot)
library(stringr)
library(kableExtra)

sex_palette <- c("#4CAF50","#F44336","#2196F3")
sex_palette2 <- c("#F44336","#2196F3")
```

## About the Data

This section is about the clients who identified as having a current mental health issue and received specialist homelessness services (SHS).

**Specialist homelessness service(s) (SHS)** is assistance provided by a specialist homelessness agency to a client aimed at responding to or preventing homelessness. The specialist homelessness services in scope for this collection include accommodation provision, assistance to sustain housing, mental health services, family/relationship assistance, disability services, drug/alcohol counselling, legal/financial services, immigration/cultural services, domestic/family violence services, other specialist services and general assistance and support.

The data used for this section came from [Australian Institute of Health and Welfare (AIHW)](https://www.aihw.gov.au/mental-health/topic-areas/specialist-homelessness-services).

```{r}
data_dir <- file.path(here::here(), "data", "AIHW")
shs2 <- readRDS(file.path(
  data_dir,
  "shs2-shs-mental-health-issue-demographic.rds"
))

# Table name: *`r shs2$table_name[1]`*. 
# 
# Table code: *`r shs2$table_code[1]`*. 
# 
# Data source: *`r shs2$source[1]`*.

# clean the variable names for visualisation
shs2$measurement <- gsub("Per cent", "Percentage",shs2$measurement)

y <- shs2 %>%
  filter(mental_health_issue == "True")
```

### By Age Group

```{r}
#| label: fig-shs-a
#| fig-cap: Recipients of SHS by Age Group, 2020-21

n_levels <- length(unique(y$client_demographics))
rmhc_blue_pal <- colorRampPalette(c("#91BFD0", "#08306B"))(n_levels)

# fig.height=10
y %>% 
  filter(group == "Age group") %>% 
  ggplot(aes(x = client_demographics, y = value, fill = client_demographics))+
  geom_col()+
  facet_wrap(~measurement,
             scales = "free_y",
             ncol = 1)+
  scale_y_continuous(labels = comma)+
  labs(x = "Age group",
       y = "Value",
       fill = "Age group")+
  theme_bw()+
  scale_fill_manual(values = rmhc_blue_pal)
```

### By Sex

```{r}
#| label: fig-shs-s
#| fig-cap: Recipients of SHS by Sex, 2020-21

y %>% 
  filter(group == "Sex") %>% 
  ggplot(aes(x = client_demographics, y = value, fill = client_demographics))+
  geom_col()+
  facet_wrap(~measurement,
             scales = "free_y")+
  scale_y_continuous(labels = comma)+
  labs(x = "Sex",
       y = "Value")+
  theme_bw()+
  theme(legend.position = "none")+
  scale_fill_manual(values = sex_palette2)
```

### Summary

SHS recipients' rate:

-   Young adults aged 15-24 constitute the highest percentage of those with current mental health issues who also receive SHS.

-   A decline in SHS recipient rates is observed as age increases.

-   Females significantly outnumber males in receiving these services.

SHS recipients' characteristics:

-   The characteristics are similar to those of the general population - age distribution is bell-shaped and 25-44.

-   The number and percentage of female recipients are much higher than males.
