
# Psychosocial Disability Support Services
```{r, message=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(forcats)
library(scales)
library(gridExtra)
library(cowplot)
library(stringr)
library(kableExtra)

sex_palette <- c("#4CAF50","#F44336","#2196F3")
sex_palette2 <- c("#F44336","#2196F3")
```


**Psychosocial disability\* support services** are critical in assisting people with psychosocial disability overcome functional limitations (for example, with communication, daily living or self-care) and facilitating full and equal participation in the community.

This section presents information on specialist disability support services provided under the National Disability Insurance Scheme (NDIS) to participants with a **primary disability\*** of psychosocial disability.

The data used for this section came from [Australian Institute of Health and Welfare (AIHW)](https://www.aihw.gov.au/mental-health/topic-areas/psychosocial-disability-support).

```{r}
data_dir <- file.path(here::here(), "data", "AIHW")
dis4 <- readRDS(file.path(
  data_dir,
  "dis4-ndis-active-by-demographic.rds"
))

dis4_sa <- dis4 %>% 
  filter(group %in% c("State_or_territory", "Age_group"))

# Table name: *`r dis4$table_name[1]`*. 
# 
# Table code: *`r dis4$table_code[1]`*. 
# 
# Data source: *`r dis4$source[1]`*.
```

## By States or Territory

```{r}
#| label: fig-dis-s
#| fig-cap: NDIS Psychosocial Participants, by States or Territory, as of Dec 2021
dis4_sa <- dis4_sa %>% 
  filter(group == "State_or_territory")

# add state levels
state_levels <- c("NSW","Vic","Qld","WA","SA","Tas","ACT","NT")
dis4_sa$participant_demographics <- factor(dis4_sa$participant_demographics,levels = state_levels)

ggplot(dis4_sa, aes(x = participant_demographics, y = value))+
geom_col()+
facet_wrap(~measurement,
           scales = "free_y",
           ncol = 1) +
scale_y_continuous(labels = comma) +
labs(x = "States or Territory",
     y = "Value")+
theme_bw()

# for analysis
# dis4_sa_a <- dis4_sa %>% 
#   filter(measurement == "Rate (per 100,000 population)")
```

## By Age

```{r}
#| label: fig-dis-a
#| fig-cap: NDIS Psychosocial Participants, by Age Group, as of Dec 2021
dis4_age <- dis4 %>% 
  filter(group == "Age_group")

n_levels <- length(unique(dis4_age$participant_demographics))
dis_blue_pal <- colorRampPalette(c("#B3E2F2", "#08306B"))(n_levels)

dis4_age %>% 
  filter(group == "Age_group") %>% 
  ggplot(aes(x = participant_demographics, y = value, fill = participant_demographics))+
  geom_col()+
  facet_wrap(~measurement,
             scales = "free_y",
             ncol = 1) +
  scale_y_continuous(labels = comma) +
  labs(x = "Age group",
       y = "Value",
       fill = "Age groups")+
  theme_bw()+
  scale_fill_manual(values = dis_blue_pal)+
  theme(legend.position = "none")

# for analysis
# dis4_sa_a2 <- dis4 %>% 
#   filter(group == "Age_group",
#          measurement == "Rate (per 100,000 population)")
```

## Summary

-   Vic and ACT have the highest rates, with approximately 250 per 100,000. Other states and territories show similar, lower rates, ranging from 170 to 190 per 100,000.

-   The rate, number, and percentage of active patients all display a similar trend across age groups, increasing progressively with age but experiencing a decline after age 65+.
