---
title: "(ABS) Persons 16–85 years, 12-month mental disorders by age and sex"
author: "Janie Chen"
format: html
---

```{r, message=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(forcats)
```

## Load data

```{r}
abs3_data_dir <- file.path(here::here(), "data", "ABS")
abs3 <- readRDS(file.path(
  abs3_data_dir,
  "abs-12-month-mental-disorders-by-age-and-sex.rds"
))
```

Table name is *`r abs3$table_name[1]`*. 
Table code is *`r abs3$table_code[1]`*. 
Data source is *`r abs3$source[1]`*.

# Population

## Global view of the population by age

### Bigger age gap groups

```{r}
ttl_pop1 <- abs3 %>% 
  filter(age_group_type == "type1",
         sex == "Persons",
         name == "Estimates")

ggplot(ttl_pop1, aes(x = disorder_types, 
                     y = value, 
                     fill = age_group)) +
  geom_bar(position="dodge", stat="identity")+
  coord_flip()+
  labs(title = "Estimate Population of Mental Disorder by Age",
       x = "Disorder types",
       y = "Population('000)",
       fill = "Age groups")+
  guides(fill = guide_legend(reverse=TRUE))
```

### Smaller age gap groups

```{r}
ttl_pop2 <- abs3 %>% 
  filter(age_group_type == "type2",
         sex == "Persons",
         name == "Estimates")

ggplot(ttl_pop2, aes(x = disorder_types, y = value, fill = age_group)) +
  geom_bar(position="dodge", stat="identity")+
  coord_flip()+
  labs(title = "Estimate Population of Mental Disorder by Age",
       x = "Disorder types",
       y = "Population('000)",
       fill = "Age groups")+
  guides(fill = guide_legend(reverse=TRUE))
```

## Global view of the population by sex

```{r}
ttl_pop_sex1 <- abs3 %>% 
  filter(sex != "Persons",
         name == "Estimates")

ggplot(ttl_pop_sex1, aes(x = disorder_types, y = value, fill = sex)) +
  geom_bar(position="dodge", stat="identity")+
  coord_flip()+
  labs(title = "Estimate Population of Mental Disorder by Sex",
       x = "Disorder types",
       y = "Population('000)",
       fill = "Sex")+
  guides(fill = guide_legend(reverse=TRUE))
```

## Population by age and sex

### Bigger age gap groups

```{r}
# Plots for bigger age groups
big_pop_sex <- abs3 %>% 
  filter(age_group_type == "type1",
         name == "Estimates",
         sex != "Persons")

ggplot(big_pop_sex, aes(x = disorder_types, y = value, fill = sex)) +
  geom_bar(position="dodge", stat="identity")+
  coord_flip()+
  labs(title = "Estimate Population of Mental Disorder by Age and Sex",
       x = "Disorder types",
       y = "Population('000)",
       fill = "Sex")+
  facet_wrap(~age_group, ncol = 1)+
  guides(fill = guide_legend(reverse=TRUE))
```

### Smaller age gap groups

```{r}
# Plots for smaller age groups
sml_pop_sex <- abs3 %>% 
  filter(age_group_type == "type2",
         name == "Estimates",
         sex != "Persons")

ggplot(sml_pop_sex, aes(x = disorder_types, y = value, fill = sex)) +
  geom_bar(position="dodge", stat="identity")+
  coord_flip()+
  labs(title = "Estimate Population of Mental Disorder by Age and Sex",
       x = "Disorder types",
       y = "Population('000)",
       fill = "Sex")+
  facet_wrap(~age_group)+
  guides(fill = guide_legend(reverse=TRUE))
```

# Proportion

## Global view of the proportion by age

### Bigger age gap groups

```{r}
ttl_pro1 <- abs3 %>% 
  filter(age_group_type == "type1",
         sex == "Persons",
         name == "Proportions")

ggplot(ttl_pro1, aes(x = disorder_types, y = value, fill = age_group)) +
  geom_bar(position="dodge", stat="identity")+
  coord_flip()+
  labs(title = "Proportion of Mental Disorder by Age",
       x = "Disorder types",
       y = "Proportion(%)",
       fill = "Age groups")+
  guides(fill = guide_legend(reverse=TRUE))
```

### Smaller age gap groups

```{r}
ttl_pro2 <- abs3 %>% 
  filter(age_group_type == "type2",
         sex == "Persons",
         name == "Proportions")

ggplot(ttl_pro2, aes(x = disorder_types, y = value, fill = age_group)) +
  geom_bar(position="dodge", stat="identity")+
  coord_flip()+
  labs(title = "Proportion of Mental Disorder by Age",
       x = "Disorder types",
       y = "Proportion(%)",
       fill = "Age groups")+
  guides(fill = guide_legend(reverse=TRUE))
```

## Global view of the proportion by sex

```{r}
ttl_pro_sex1 <- abs3 %>% 
  filter(sex != "Persons",
         name == "Proportions")

ggplot(ttl_pro_sex1, aes(x = disorder_types, y = value, fill = sex)) +
  geom_bar(position="dodge", stat="identity")+
  coord_flip()+
  labs(title = "Proportion of Mental Disorder by Sex",
       x = "Disorder types",
       y = "Proportion(%)",
       fill = "Sex")+
  guides(fill = guide_legend(reverse=TRUE))
```

## Proportion by age and sex

### Bigger age gap groups

```{r}
# Plots for bigger age groups
big_pro_sex <- abs3 %>% 
  filter(age_group_type == "type1",
         name == "Proportions",
         sex != "Persons")

ggplot(big_pro_sex, aes(x = disorder_types, y = value, fill = sex)) +
  geom_bar(position="dodge", stat="identity")+
  coord_flip()+
  labs(title = "Proportion of Mental Disorder by Age and Sex",
       x = "Disorder types",
       y = "Proportion(%)",
       fill = "Sex")+
  facet_wrap(~age_group, ncol = 1)+
  guides(fill = guide_legend(reverse=TRUE))
```

### Smaller age gap groups

```{r}
# Plots for smaller age groups
sml_pop_sex <- abs3 %>% 
  filter(age_group_type == "type2",
         name == "Proportions",
         sex != "Persons")

ggplot(sml_pop_sex, aes(x = disorder_types, y = value, fill = sex)) +
  geom_bar(position="dodge", stat="identity")+
  coord_flip()+
  labs(title = "Proportion of Mental Disorder by Age and Sex",
       x = "Disorder types",
       y = "Proportion(%)",
       fill = "Sex")+
  facet_wrap(~age_group)+
  guides(fill = guide_legend(reverse=TRUE))
```