---
title: "Productivity Commission Report on Government Services 2023"
subtitle: "(ROGS 13A.45 and 46)"
author: "Janie Chen"
format: html
editor_options: 
  chunk_output_type: inline
execute:
  warning: FALSE
  echo: TRUE
---


```{r, message=FALSE}
library(ggplot2)
library(dplyr)
library(scales)
library(tidyverse)
library(tidyr)
library(knitr)
library(kableExtra)

sex_palette <- c("#4CAF50","#F44336","#2196F3")
```
# Tab 45

‘Prevalence of mental illness’ is defined as the proportion of the total population who have a mental illness.

A low or decreasing prevalence of mental illness can indicate that measures to prevent mental illness have been effective.

Due to COVID-related difficulties with in-person data collection during 2021, the National Study of Mental Health and Wellbeing (NSMHW) was conducted in two parts from 2020 to 2022. As a result, only national totals are available for reporting in 2023. *These data should not be compared to the 2007 survey due to a smaller sample size in 2020-2021*. State and territory data are expected to be available for the 2024 Report.

```{r}
data_dir <- file.path(here::here(), "data", "ROGS")
rogs45 <- readRDS(file.path(
  data_dir,
  "prevalence-among-adults-by-sex.rds"
)) %>% 
  select(-c(Data_Source, Description6))
```

```{r}
data_dir <- file.path(here::here(), "data", "ROGS")
rogs46 <- readRDS(file.path(
  data_dir,
  "prevalence-among-adults-by-age.rds"
)) %>% 
  select(-c(Data_Source, Description4))
```

# Prevalence of mental illness for all population by sex and age for 2020-21 and 2007

## Disorder Timeframe by Sex

```{r}
# Data wrangling 
tab <- rogs45 %>% 
  mutate(Aust = as.numeric(Aust)) %>% 
  filter(Uncertainty == "") %>% 
  select(-c(Uncertainty,Unit)) %>% 
  rename(disorder_timeframe = Description2,
    disorder_group = Description3,
    specific_disorder = Description4)

tab$disorder_timeframe <- str_replace_all(tab$disorder_timeframe, 
                                            "Lifetime mental disorder; with no 12-month symptoms", 
                                            "Lifetime mental disorder;\nwith no 12-month symptoms")

# plot
disorder_timeframe <- tab %>% 
  filter(disorder_group == "" | str_detect(disorder_group, "Total"))

ggplot(disorder_timeframe, aes(x = disorder_timeframe, y = Aust, fill = Sex))+
  geom_bar(position="dodge", stat="identity")+
  facet_wrap(.~Year,ncol=1)+
  coord_flip()+
  labs(title = "Disorder timeframe by sex for 2020-21 and 2007",
       x = "Disorder timeframe",
       y = "Percentage for all population")+
  theme_bw()
```

## Disorder Timeframe by Age

```{r}
# Data wrangling 
tab_46 <- rogs46 %>% 
  mutate(Aust = as.numeric(Aust)) %>% 
  filter(Uncertainty == "") %>% 
  select(-c(Uncertainty,Unit)) %>% 
  rename(disorder_timeframe = Description2,
    disorder_group = Description3)

tab_46$disorder_timeframe <- str_replace_all(tab_46$disorder_timeframe, 
                                            "Lifetime mental disorder; with no 12-month symptoms", 
                                            "Lifetime mental disorder;\nwith no 12-month symptoms")

tab_46$Age <- gsub(" years old","",tab_46$Age)

# plot
disorder_timeframe_46 <- tab_46 %>% 
  filter(disorder_group == "" | str_detect(disorder_group, "Total"))

ggplot(disorder_timeframe_46, aes(x = Age, y = Aust, fill = Year))+
  geom_bar(position="dodge", stat="identity")+
  facet_wrap(.~disorder_timeframe)+
  scale_fill_manual(values = c("#0C356A","#279EFF"))+
  coord_flip()+
  labs(title = "Disorder timeframe by age for 2020-21 and 2007",
       y = "Percentage for all population")+
  theme_bw()+
  guides(fill = guide_legend(reverse=TRUE))
```


## Disorder Group by Sex

```{r}
disorder_group <- tab %>% 
  filter(disorder_timeframe == "Any 12-month mental disorder",
         !str_detect(disorder_group, "Total")) %>% 
  select(-disorder_timeframe)

disorder_group_ttl <- disorder_group %>%
  filter(specific_disorder=="") 
disorder_group_ttl$disorder_group <- gsub("Any ","",disorder_group_ttl$disorder_group)
disorder_group_ttl$disorder_group <- tools::toTitleCase(disorder_group_ttl$disorder_group)

ggplot(disorder_group_ttl, aes(x = disorder_group, y = Aust, fill = Sex))+
  geom_bar(position="dodge", stat="identity")+
  scale_fill_manual(values = sex_palette)+
  facet_wrap(.~Year, ncol = 1)+
  coord_flip()+
  labs(title = "Disorder group by sex for 2020-21 and 2007",
       x = "Disorder Group",
       y = "Percentage for all population")+
  theme_bw()
```


## Disorder Group by Age

```{r}
disorder_group_46 <- tab_46 %>% 
  filter(disorder_timeframe == "Any 12-month mental disorder",
         !str_detect(disorder_group, "Total")) %>% 
  select(-disorder_timeframe)

ggplot(disorder_group_46, aes(x = Age, y = Aust, fill = Year))+
  geom_bar(position="dodge", stat="identity")+
  facet_wrap(.~disorder_group)+
  scale_fill_manual(values = c("#0C356A","#279EFF", "#40F8FF"))+
  coord_flip()+
  labs(title = "Disorder group by age for 2020-21 and 2007",
       y = "Percentage of all population")+
  theme_bw()+
  guides(fill = guide_legend(reverse=TRUE))
```


## Specific Disorder Types Distribution 

```{r}
specific_disorder_tab <- disorder_group %>% 
  filter(specific_disorder!="") %>% 
  pivot_wider(names_from = c(Sex,Year),
              values_from = Aust)

specific_disorder_tab %>% 
  select(-disorder_group) %>% 
  kable(format = "html",
        col.names = c("Specific Disorder",rep(c("Male(%)","Female(%)","All(%)"),2)),
        caption = "Specific disorder types by sex for 2020-21 and 2007",
      ) %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>% 
  kable_classic(full_width = F, html_font = "Cambria") %>% 
  add_header_above(c(" "=1, "2020-21"=3, "2007"=3), bold = TRUE) %>% 
  column_spec(c(1,4), extra_css = "border-right: 2px solid grey;") %>% 
  row_spec(0, bold = TRUE) %>% 
  pack_rows("Anxiety disorders",1,6) %>% 
  pack_rows("Affective Disorders",7,9) %>% 
  pack_rows("Substance use disorders",10,12) %>% 
  column_spec(c(2,5), color = "#2196F3") %>% 
  column_spec(c(3,6), color = "#F44336")
```

