---
title: "Residential mental health care services"
subtitle: "(AIHW RMHC.3,4,6,7)"
author: "Janie Chen"
format: html
editor_options: 
  chunk_output_type: console
execute:
  warning: FALSE
  echo: TRUE
toc: true
toc-floating: true
---

```{r, message=FALSE}
library(ggplot2)
library(dplyr)
library(scales)
```

# Demographic of people accessing residential mental health care

## Load data

```{r}
data_dir <- file.path(here::here(), "data", "AIHW")
rmhc3 <- readRDS(file.path(
  data_dir,
  "rmhc3-ppl-accessing-rmhc.rds"
))
```

Table name: *`r rmhc3$table_name[1]`*. 

Table code: *`r rmhc3$table_code[1]`*. 

Data source: *`r rmhc3$source[1]`*.


## Age distribution

```{r}
rmhc3 %>% 
  filter(demographic == "Age_group") %>% 
  ggplot(aes(x = demographic_variable, y = value))+
  geom_col()+
  facet_wrap(~measurement,
             scales = "free_y",
             ncol = 1)+
  scale_y_continuous(labels = comma)+
  labs(x = "Age Groups",
       y = "Value",
       title = "Age distribution of people accessing residential mental health care")
```

## Sex distribution

```{r}
rmhc3 %>% 
  filter(demographic == "Sex") %>% 
  ggplot(aes(x = demographic_variable, y = value))+
  geom_col()+
  facet_wrap(~measurement,
             scales = "free_y",
             ncol = 3)+
  scale_y_continuous(labels = comma)+
  labs(x = "Sex",
       y = "Value",
       title = "Sex distribution of people accessing residential mental health care")
```

# The rate of residential mental health care episodes

## Load data

```{r}
data_dir <- file.path(here::here(), "data", "AIHW")
rmhc4 <- readRDS(file.path(
  data_dir,
  "rmhc4-rmhc-by-sex-age-states.rds"
))
```

Table name: *`r rmhc4$table_name[1]`*. 

Table code: *`r rmhc4$table_code[1]`*. 

Data source: *`r rmhc4$source[1]`*.

```{r}
data_dir <- file.path(here::here(), "data", "AIHW")
rmhc7 <- readRDS(file.path(
  data_dir,
  "rmhc7-rmhc-ts-by-age-sex.rds"
))

rmhc7_rsa <- rmhc7 %>% 
  filter(count == "Rate_per_10_000_population",
         demographic %in% c("Sex", "Age_group"))
rmhc7_rsa$year <- gsub("–\\d{2}", "", rmhc7_rsa$year)
rmhc7_rsa$year <- as.numeric(rmhc7_rsa$year)
```

Table name: *`r rmhc7$table_name[1]`*. 

Table code: *`r rmhc7$table_code[1]`*. 

Data source: *`r rmhc7$source[1]`*.


## By sex and age

```{r}
sex_palette <- c("#4CAF50","#F44336","#2196F3")

rmhc4 %>% 
  filter(count != "Number",
         states_and_territories == "Total") %>% 
  ggplot(aes(x = age_group, y = value, fill = sex))+
  geom_bar(position="dodge", stat="identity")+
  scale_fill_manual(values = sex_palette)+
  labs(x = "Age groups",
       y = "Rate (per 10,000 population)",
       title = "Total episodes rate by age and sex")
```


## By sex, states and territories

```{r}
rmhc4 %>% 
  filter(count != "Number",
         states_and_territories != "Total",
         age_group != "Total") %>% 
  ggplot(aes(x = states_and_territories, y = value, fill = sex))+
  geom_bar(position="dodge", stat="identity")+
  scale_fill_manual(values = sex_palette)+
  labs(x = "States and territories",
       y = "Rate (per 10,000 population)",
       title = "Episodes rate by states and territories")
```


## By sex, age, states and territories

```{r}
rmhc4 %>% 
  filter(count != "Number",
         age_group != "Total",
         states_and_territories != "Total") %>% 
  ggplot(aes(x = age_group, y = value, fill = sex))+
  geom_bar(position="dodge", stat="identity")+
  scale_fill_manual(values = sex_palette)+
  facet_wrap(~states_and_territories,
             ncol = 2)+
  labs(x = "Age group",
       y = "Rate (per 10,000 population)",
       title = "Episodes rate by sex, age, states and territories")
```


## Changes over the year by age group

```{r}
rmhc7_rsa %>% 
  filter(demographic == "Age_group") %>% 
  ggplot(aes(x = year, y = value)) +
  geom_line()+
  facet_wrap(~demographic_variable)+
  labs(x = "Year",
       y = "Rate (per 10,000 population)",
       title = "Episodes rate changes by age over year 2005–06 to 2020–21")
```

## Changes over the year by sex

```{r}
rmhc7_rsa %>% 
  filter(demographic == "Sex") %>% 
  ggplot(aes(x = year, y = value)) +
  geom_line()+
  facet_wrap(~demographic_variable)+
  labs(x = "Year",
       y = "Rate (per 10,000 population)",
       title = "Episodes rate changes by sex over year 2005–06 to 2020–21")
```


Note: didn't analyse episodes *numbers*, will add it on if needed.

# Residential mental health care episodes per 10,000 population

## Load data

```{r}
data_dir <- file.path(here::here(), "data", "AIHW")
rmhc6 <- readRDS(file.path(
  data_dir,
  "rmhc6-rmhc-per-10000-pop-by-sex-n-age.rds"
))

rmhc6$year <- gsub("–\\d{2}", "", rmhc6$year)
rmhc6$year <- as.numeric(rmhc6$year)
```

Table name: *`r rmhc6$table_name[1]`*. 

Table code: *`r rmhc6$table_code[1]`*. 

Data source: *`r rmhc6$source[1]`*.

## Average Annual Change

```{r}
sex_palette <- c("#4CAF50","#F44336","#2196F3")

ap <- rmhc6 %>% 
  group_by(age_group,sex) %>% 
  summarise(avg = mean(average_annual_change_per_cent_2016_17_to_2020_21))


a_hline <- ap %>% filter(age_group == "Total", sex == "All_persons") %>% pull(avg)
f_hline <- ap %>% filter(age_group == "Total", sex == "Females") %>% pull(avg)
m_hline <- ap %>% filter(age_group == "Total", sex == "Males") %>% pull(avg)

ggplot(ap, 
       aes(x = age_group,
           y = avg,
           fill = sex)) +
  geom_bar(position="dodge", stat="identity")+
  scale_fill_manual(values = sex_palette)+
  labs(x = "Age groups",
       y = "Average annual change (per cent)",
       title = "Average Annual Change Percent by Sex and Age group, 2016–17 to 2020–21")+
  geom_hline(yintercept = a_hline, linetype = "dashed", color = "#4CAF50")+
  geom_hline(yintercept = f_hline, linetype = "dashed", color = "#F44336")+
  geom_hline(yintercept = m_hline, linetype = "dashed", color = "#2196F3")+
  geom_text(aes(x = "85+", y = a_hline, label = "avg. All persons"), hjust = 0.8, vjust = -0.5, size = 3, color = "#4CAF50") +
  geom_text(aes(x = "85+", y = f_hline, label = "avg. Females"), hjust = 0.8, vjust = -0.5, size = 3, color = "#F44336") +
  geom_text(aes(x = "85+", y = m_hline, label = "avg. Males"), hjust = 0.8, vjust = -0.5, size = 3, color = "#2196F3")
  
```


## Changes Over Time by Age Group 

### Male mental health care episodes changes over the year

```{r}
mts <- rmhc6 %>% 
  filter(sex == "Males")

ggplot(mts, aes(x = year, y = episodes_per_10_000_pop)) +
  geom_line()+
  facet_wrap(~age_group)+
  labs(x = "Year",
       y = "Episodes (per 10,000 population)",
       title = "Male Episodes Change by Age Groups")
```

### Female mental health care episodes changes over the year

```{r}
fts <- rmhc6 %>% 
  filter(sex == "Females")

ggplot(fts, aes(x = year, y = episodes_per_10_000_pop)) +
  geom_line()+
  facet_wrap(~age_group)+
  labs(x = "Year",
       y = "Episodes (per 10,000 population)",
       title = "Female Episodes Change by Age Groupss")
```

